Short and Mechanized Logical Relation for Dependent Type Theory

### 2024-02-22 paper/paper.tex
minor wording in intro

\section{Introduction}
In this paper, we present a \emph{short}, \emph{mechanized} proof of logical
consistency for \lang{}, a Martin-Löf style dependent type theory with large
eliminations, a full predicative universe hierarchy, an intensional identity
type and a boolean base type, both of which are equipped with dependent
elimination forms.

Our goal with this work is to demonstrate how the proof technique of
\emph{logical relations} can be applied in dependently-typed setting.
Tutorial material on this topic~\cite{lr-tutorials} is primarily focussed on
languages with simple or polymorphic types. In that context, logical relations
can be define as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation.

Neither of these techniques make sense for predicative dependent type theory,
so a young researcher might be excused for thinking that logical
relations-based arguments are not available.  \scw{Well, actually, we did use
  a step-indexed relation for Trellys\ldots}\yl{The core calculus of Trellys' logical fragment
  is simple type + identity so I don't think that counts..} But this is not the case. Recent
authors have developed tour-de-force mechanizations for the metatheory of
modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have used logical-relations-based reasoning as part of their
arguments. However, at the same time, these definitions have been overshadowed
by their magnitude of the surrounding developments. Because these proofs show
diverse results about real systems and algorithms, these developments range in
size from 20,000 to 400,000 lines of code, and the techniques are not
accessible to a casual reader.

>>>>

\section{Introduction}
In this paper, we present a \emph{short} and \emph{mechanized} proof of
logical consistency for \lang{}, a dependent type theory with a full
predicative universe hierarchy, large eliminations, an intensional identity
type, a boolean base type, and dependent elimination forms.

Our goal with this work is to demonstrate how the proof technique of
\emph{logical relations} can be applied to reason about dependent type theories.
Tutorial material on this topic~\cite{lr-tutorials} is primarily focussed on
systems with simple or polymorphic types. In that context, logical relations
can be define as simple recursive functions over the structure of types, or
(in the case of recursive types) defined over the evaluation steps of the
computation.

Neither of these techniques make sense for predicative dependent type theory,
so a young researcher might be excused for thinking that proofs that use
logical relations are not applicable.  \scw{Well, actually, we did use a
  step-indexed relation for Trellys\ldots}\yl{The core calculus of Trellys'
  logical fragment is simple type + identity so I don't think that counts..}
But this is not the case. Recent authors have developed tour-de-force
mechanizations for the metatheory of modern proof
assistants~\citep{nbeincoq,decagda,martin-lof-a-la-coq,anand2014towards}, and
have relied on logical relations defined as part of their
developments. However, because these proofs show diverse results about real
systems and algorithms, these developments range in size from 20,000 to
400,000 lines of code, these definitions are easy to miss and difficult to
isolate from their surrounding contexts. As a result, they are not accessible
to casual readers.
